---
output: github_document

---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "inst/README-"
)
```
[![](http://www.r-pkg.org/badges/version/repo)](https://cran.r-project.org/package=repo) <sup><sub>Master: </sub></sup>[![Travis-CI Build Status](https://travis-ci.org/franapoli/repo.svg?branch=master)](https://travis-ci.org/franapoli/repo) <sup><sub>Dev: </sub></sup>[![Travis-CI Build Status](https://travis-ci.org/franapoli/repo.svg?branch=dev)](https://travis-ci.org/franapoli/repo)


## Repo

Repo is a data-centered data flow manager. It allows to store R data
files in a central local repository, together with tags, annotations,
provenance and dependence information. Any saved object can then be
easily located and loaded through the repo interface.

A [paper about Repo](http://rdcu.be/pklt) has been published in BMC
Bioinformatics.

Latest news are found in the NEWS.md file of the "Untested" branch.


## Minimal example

Creating a dummy repository under the R temporary folder (skipping
confirmation):

```{r}
library(repo)
rp <- repo_open(tempdir(), force=T)
```

Storing data. In this case, just item values and names are specified:

```{r}
God <- Inf
rp$put(God)          ## item name inferred from variable name
rp$put(0, "user")    ## item name specified
```

More data with specified dependencies:

```{r}
rp$put(pi, "The Pi costant", depends="God")
rp$put(1:10, "r", depends="user")
```

Loading items from the repository on the fly using names:

```{r}
diam <- 2 * rp$get("r")
circum <- 2 * rp$get("The Pi costant") * rp$get("r")
area <- rp$get("The Pi costant") * rp$get("r") ^ 2
```

Storing more data with verbose descriptions:

```{r}
rp$put(diam, "diameters", "These are the diameters", depends = "r")
rp$put(circum, "circumferences", "These are the circumferences",
       depends = c("The Pi costant", "r"))
rp$put(area, "areas", "These are the areas",
       depends = c("The Pi costant", "r"))
```

Showing repository contents:

```{r}
print(rp)
```
```{r}
rp$info()
```
```{r}
rp$info("areas")
```

Visualizing dependencies:

```{r, depgraph}
rp$dependencies()
```

Manual acces to stored data:

```{r}
fpath <- rp$attr("r", "path")
readRDS(fpath)
```

## Development branches

+ [Master](https://github.com/franapoli/repo/tree/master): stable major
releases, usually in sync with lastest CRAN version.

+ [Dev](https://github.com/franapoli/repo/tree/dev): minor releases
passing automatic checks.

+ [Untested](https://github.com/franapoli/repo/tree/untested): in
progress versions and prototype code, not necessarily working.


## Manuals

Besides inline help, two documents are available as introductory
material:

+ [A paper published on BMC Bioinformatics](http://rdcu.be/pklt)

+ [An introductory
vignette](https://rawgit.com/franapoli/repo/gh-pages/index.html).


## Download and Installation

Repo is on CRAN and can be installed from within R as follows:

    > install.packages("repo")
    
Latest stable release can be downloaded from Github at
[https://github.com/franapoli/repo](https://www.github.com/franapoli/repo/).
Repo can then be installed from the downloaded sources as follows:

    > install.packages("path-to-downloaded-source", repos=NULL)

`devtools` users can install directly from github as follows:

    > install_github("franapoli/repo", ref="dev")


