---
title: "The *Repo* R Data Manager"
date:
author: "Francesco Napolitano"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Repo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=7, fig.height=7, comment="")
```

## Introduction

This is a getting-started guide for the **Repo** R package, which
implements an R objects repository manager.  It is a **data-centered
data flow manager**.

The Repo package builds one (or more) centralized local repository
where R objects are stored together with corresponding annotations,
tags, dependency notes, provenance traces, source code. Once a
repository has been populated, stored objects can be easily searched,
navigated, edited, imported/exported. Annotations can be exploited to
reconstruct data flows and perform typical pipeline management
operations.

Additional information can be found in the paper: Napolitano,
F. [*repo: an R package for data-centered management of bioinformatic
pipelines*](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/s12859-017-1510-6). **BMC
Bioinformatics** 18, 112 (2017).

Repo latest version can be found at: https://github.com/franapoli/repo

Repo is also on CRAN at: https://cran.r-project.org/package=repo

## Preparation

The following command creates a new repository in a temporary path
(the default would be "~/.R_repo"). The same function opens existing
repositories. The variable `rp` will be used as the main interface to
the repository throughout this guide.

```{r}
library(repo)
rp <- repo_open(tempdir(), force=T)
```

This document is produced by a script named `index.Rmd`. The script
itself can be added to the repository and newly created resources
annotated as being produced by it. The annotation is made automatic
using the `options` command.


```{r}
rp$attach("index.Rmd", "Source code for Repo vignette")
rp$options(src="index.Rmd")
```


## Populating the repository

Here is a normalized version of the *Iris* dataset to be stored in the
repository:

```{r, eval=F}
myiris <- scale(as.matrix(iris[,1:4]))
```

The shortest way to permanently store the `myiris` object in the
repository is simply:

```{r, eval=F}
rp$put(myiris)
```

However, richer annotation is possible, for example:

```{r}
## chunk "myiris" {
myiris <- scale(as.matrix(iris[,1:4]))

rp$put(
    obj = myiris,
    name = "myiris", 
    description = paste(
        "A normalized version of the iris dataset coming with R.",
        "Normalization is made with the scale function",
        "with default parameters."
    ),
    tags = c("dataset", "iris", "repodemo")
)
## }
```

The call provides the data to be stored (`obj`), an identifier
(`name`), a longer `description`, a list of `tags`.

The comment lines (`## chunk "myiris" {` and `## }`) have a special
meaning: they associate the corresponding code to the resource. The
code can be showed as follows:

```{r}
rp$chunk("myiris")
```

The code associated with an item should take care of building and
storing it. The *build* command executes the code in the current
environment. It can automatically build dependencies, too.

```{r}
rp$rm("myiris")
rp$build("myiris", "index.Rmd")
```


In this example, the Iris class annotation will be stored
separately:

```{r}
rp$put(iris$Species, "irisLabels",
		     tags = c("labels", "iris", "repodemo"))
```

### Attaching figures

The following code produces a 2D visualization of the Iris data
and shows it:

```{r}
irispca <- princomp(myiris)
iris2d <- irispca$scores[,c(1,2)]
plot(iris2d, main="2D visualization of the Iris dataset",
     col=rp$get("irisLabels"))
```

Note that `irisLabels` is loaded on the fly from the repository.

It would be nice to store the figure itself in the repo together with
the Iris data. This is done using the `attach` method, which stores
any file in the repo as is (as opposed to R objects), plus
annotations. Two parameters differ from `put`:

* **filepath** Instead of an identifier, `attach` takes a file name
(with path). The file name will be also the item identifier.

* **to** This optional parameter tells Repo which item the new one is
attached to. Can be empty.


```{r}
fpath <- file.path(rp$root(), "iris2D.pdf")
pdf(fpath)
plot(iris2d, main="2D visualization of the Iris dataset",
     col=rp$get("irisLabels"))
invisible(dev.off())
rp$attach(fpath, "Iris 2D visualization obtained with PCA.",
            c("visualization", "iris", "repodemo"),
              to="myiris")
```

The attached PDF can be accessed using an external PDF viewer directly
from within Repo through the `sys` command. On a Linux system, this
command runs the Evince document viewer and shows `iris2D.pdf`:

```{r, eval=FALSE}
rp$sys("iris2D.pdf", "evince")
```

The following code makes a clustering of the Iris data and stores it
in the repository. There is one parameter to note:

* **depends** Tells Repo that, in order to compute the `kiris`
variable, `myiris` is necessary. (This information is used by `build`
to build dependencies and by `dependencies` to show them).

```{r}
kiris <- kmeans(myiris, 5)$cluster
rp$put(kiris, "iris_5clu", "Kmeans clustering of the Iris data, k=5.",
         c("metadata", "iris", "kmeans", "clustering", "repodemo"),
           depends="myiris")
```

The following shows what the clustering looks like. The figure
will be attached to the repository as well.

```{r}
plot(iris2d, main="Iris dataset kmeans clustering", col=kiris)
```

```{r}
fpath <- file.path(rp$root(), "iris2Dclu.pdf")
pdf(fpath)
plot(iris2d, main="Iris dataset kmeans clustering", col=kiris)
invisible(dev.off())
rp$attach(fpath, "Iris K-means clustering.",
	c("visualization", "iris", "clustering", "kmeans", "repodemo"),
	 		   to="iris_5clu")
```

Finally, a contingency table of the Iris classes versus clusters is
computed below. The special tag *hide* prevents an item from being
shown unless explicitly requested.


```{r}
res <- table(rp$get("irisLabels"), kiris)
rp$put(res, "iris_cluVsSpecies",
         paste("Contingency table of the kmeans clustering versus the",
               "original labels of the Iris dataset."),
         c("result", "iris","validation", "clustering", "repodemo", "hide"),
         src="index.Rmd", depends=c("myiris", "irisLabels", "iris_5clu"))
```

## Looking at the repository

The `info` command summarizes some information about a repository:


```{r}
rp$info()
```

The Repo library supports an S3 `print` method that shows the contents
of the repository. All non-hidden items will be shown, together with
some details, which by defaults are: name, dimensions, size.


```{r}
rp ## resolves to print(rp)
```

Hidden items are... hidden. The following will show them too:


```{r}
print(rp, all=T)
```

Items can also be filtered. With the following call, only
items tagged with "clustering" will be shown:


```{r}
print(rp, tags="clustering", all=T)
```

`print` can show information selectively. This command shows tags and
size on disk:


```{r}
rp$print(show="st")
```

The `find` command will match a search string against all item fields
in the repository:

```{r}
rp$find("clu", all=T)
```


It is also possible to obtain a visual synthetic summary of the
repository by using the `pies` command:

```{r}
rp$pies()
```

Finally, the `check` command runs an integrity check verifying that
the stored data has not been modified/corrupted. The command will also
check the presence of extraneous (not indexed) files. Since the `rp`
repository was created in a temporary directory, a few extraneous
files will pop up.

```{r}
rp$check()
```


### Showing dependencies

In Repo, the relations "generated by", "attached to" and "dependent
on" are summarized in a *dependency graph*. The formal representation
of the graph is a matrix, in which the entry (i,j) represent a
relation from i to j of type 1, 2 or 3 (*dependency*, *attachment* or
*generation*). Here's how it looks like:


```{r}
depgraph <- rp$dependencies(plot=F)
library(knitr)
kable(depgraph)
```

Omitting the `plot=F` parameter, the *dependencies* method will
plot the dependency graph. This plot requires the *igraph* library.


```{r}
if(require("igraph", NULL, T, F))
     rp$dependencies()
```

The three types of edges can be shown selectively, so here's how the
graph looks like without the "generated" edges:


```{r}
if(require("igraph"))
    rp$dependencies(generated=F)
```

## Accessing items in the repo

The `get` command is used to retrieve items from a repository. In the
following the variable `myiris` is loaded into the variable `x` in the
current environment.


```{r}
x <- rp$get("myiris")
```

An even simpler command is `load`, which uses the item name also as
variable name:

```{r}
rm("myiris")
rp$load("myiris")
"myiris" %in% ls()
```

The `info` command can provide addtional information about an entry:


```{r}
rp$info("myiris")
```

## Item versions, temporary items, remote contents

There are actually 3 different ways of adding an object to a
repository:

* Add a new object (`rp$put`)
* Overwrite an existing object (`rp$put(replace=T)`)
* Add a new version of an existing object (`rp$put(replace="addversion")`)

Plus, item contents for an existing entry can be dowloaded if an URL
is provided with it (`rp$pull`).

### Versioning

The K-means algorithm will likely provide different solutions over
multiple runs. Alternative solutions can be stored as new versions of
the `iris_5clu` item as fllows:


```{r}
kiris2 <- kmeans(myiris, 5)$cluster
rp$put(kiris2, "iris_5clu",
         "Kmeans clustering of the Iris data, k=5. Today's version!",
           depends="myiris", replace="addversion")
```

* **addversion** when replace is set to "addversion", Repo will add a
new version of an existing object. The new object will replace the old
one and the old one will be renamed adding the suffix "#N", with N
being an incremental integer.

The new repository looks like the old one:


```{r}
rp
```

Except that `iris_5clu` is actually the one just put (look at the
description):


```{r}
rp$info("iris_5clu")
```

The old one has been renamed and hidden:

```{r}
rp$info("iris_5clu#1")
```

### Caching

It is also possible to use the repository for caching purposes by
using the `lazydo` command. It will run an expression and store the
results. When the same expression is run again, the results will be
loaded from the repository instead of being built again.

```{r}

## First run
system.time(rp$lazydo(
    {
	Sys.sleep(.5)
	result <- "This took half a second to compute"
    }
))

## Second run
system.time(rp$lazydo(
    {
	Sys.sleep(.5)
	result <- "This took half a second to compute"
    }
))
	

```

### Pulling

Existing items can feature an *URL* property. The `pull` function is
meant to update item contents by downloading them from the
Internet. This allows for the distribution of "stub" repositories
containing all items information without the actual data. The
following code creates an item provided with a remote URL. A call to
`pull` overwrites the stub local content with the remote content.

```{r}
rp$put("Local content", "item1",
	"This points to big data you may want to download",
	"tag", URL="http://exampleURL/repo")
print(rp$get("item1"))
```

```{r, eval=F}
rp$pull("item1", replace=T)
```

```{r, include=F}
rp$set("item1", obj="Remote content")
```

```{r}
print(rp$get("item1"))
```


## Handlers

The `handlers` method returns a list of functions by the same names of
the items in the repo. Each of these functions can call Repo methods
(`get` by default) on the corresponding items. In this way all item
names are loaded, which may be usefule for example to exploit
auto-completion features of the editor.


```{r}
h <- rp$handlers()
names(h)
```

Handlers call `get` by default:


```{r}
print(h$iris_cluVsSpecies())
```

The `tag` command (not yet described)  adds a tag to an item:


```{r}
h$iris_cluVsSpecies("tag", "onenewtag")
h$iris_cluVsSpecies("info")
```

One may want to open a repo directly with:


```{r}
h <- repo_open(rp$root())$handlers()
```


In that case, the handler to the repo itself will come handy:


```{r}
h$repo
```

If items are removed or added, handlers may need a refresh:


```{r}
h <- h$repo$handlers()
```


## Further documentation

The repo manual starts at:


```{r, eval=FALSE}
help(repo)
```

In order to get help on the function "func", try the following:


```{r, eval=FALSE}
help(repo_func)
```


```{r, include=F}
unlink(rp$root(), recursive=T)
```

<hr/><small><i>Based on Repo build  `r packageVersion("repo")`</i></small>

