---
title: "Getting started with the *repo* R Data Manager"
date:
author: "Francesco Napolitano"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with repo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(fig.width=7, fig.height=7, comment="")
```
## Introduction

This is a getting-started guide to the *repo* R package, which
implements an R objects repository manager. The motivations behind
the development of *repo* concern both scientific considerations,
as a tool to improve reproducibility in computational research, and
practical, as a solution to everyday problems of people working in
the same fields.

Data analysis is a also a data multiplication activity. Starting
from an initial resource, derived ones are built, each of which can
be used in turn to build more processed data. After working for a
while on data analysis projects, simple repetitive tasks such as
locating a resource built in the past (what was its name? in which
directory was it stored? for which project? on which machine?) or
remembering how it had been made (which model?  which parameters?
which script?) can quickly become daunting.

The *repo* package tries to help with such everyday issues. It
builds one (or more) centralized local repository where R objects
are stored together with corresponding annotations, tags,
dependency notes, provenance traces. Once a repository has been
populated, the user is presented with a structured collection of
all the stored resources, which can be easily explored, navigated,
modified, imported and exported.

Main features include:

* Store R objects in a central repository
* Retrieve any stored object simply by its name
* Store object description, tags, provenance trace
* Search items by names or tags
* Show details repository-wise or item-wise
* Build and plot items dependency trees
* Attach regular files to repo items
* Run system commands on stored attachments
* Quickly store temporary data
* Store multiple versions of the same resource
* Concurrent access to the same repository
* Copy items between repositories
* Export items back to regular R data files
* Check repository integrity through MD5 checksums

What follows is a walk-through aimed at quickly introducing to most
of the *repo* features.

## Preparation

First of all, the following will enable repo:


```{r}
library(repo)
```

The file name of a script is a good provenance trace for all the
data produced by it. The R function *normalizePath* can be used to
obtain the full path of the file. However, for this guide I won't
include paths from my hard drive:


```{r}
src <- "repodemo.R"
```

The following command creates a new repository below a temporary
path, which is ok for this demo. By default the repository would
instead be created under "~/.R_repo". The same function opens the
repository if already exists. To avoid confirmation requests during
the generation of this document, they're turned off through the
*force* parameter. The variable *repo* will be used as the main
interface to the repository throughout this guide.


```{r}
repo <- repo_open(tempdir(), force=T)
```

## Populating the repository

In this section, some data will be stored inside the repository
just created, namely: a normalized version of the Iris
dataset; a dimensionality-reduced and a clustering of the same
data; some visualizations.


### Storing data in the repository

The Iris dataset is readily found in the *datasets* R
package. However, here a normalized version of it will be stored in
the repo. Here's the data to store:


```{r}
myiris <- scale(as.matrix(iris[,1:4]))
```

A call to *repo$put* will provide the following parameters to
properly store the object in the repository:

* **obj** The variable containing the object to store: *myiris*.

* **name** An identifier for the variable. At the moment any string is
accepted, but this may change in favor of sanitized names. Best
practice is to use a short name without special characters.

* **description** A verbose description of the variable contents.

* **tags** A set of tags. Tags are used to subset the repo for many
purposes. For example in order to be able to select only items
produced by this script, a specific tag will be added accordingly
(*repodemo*).

* **src** Name of the file containing the source code that
generated the object being *put*. This part of the object's
"provenance trace".

* **replace** A logical that tells *repo* whether to overwrite an
existing entry by the same name. Setting it to *true* allows to run
a script over without interruptions.

Thus, the following call to the *put* method will store the
contents of *myiris* in the repo. The data will be actually saved
in the repo root in RDS format.


```{r}
repo$put(
    obj = myiris,    
    name = "myiris",    
    description = paste(
        "A normalized version of the iris dataset coming with R.",
        "Normalization is made with the scale function",
        "with default parameters."
    ),
    tags = c("dataset", "iris", "repodemo"),    
    src = src,    
    replace=T
)
```

In this example, the class annotation will be stored
separately. Here's a more compact call to *put*:


```{r}
repo$put(iris$Species, "irisLabels", "The Iris class lables.",
         c("labels", "iris", "repodemo"), src, replace=T)
```

### Attaching visualizations

The Iris dataset is 4D and PCA can be used to produce a 2D
visualization. The following code produces such visualization
and shows it:


```{r}
irispca <- princomp(myiris)        
iris2d <- irispca$scores[,c(1,2)]
plot(iris2d, main="2D visualization of the Iris dataset",
     col=repo$get("irisLabels"))
```

Note that *irisLabels* is taken on the fly from the repo. This
prevents the creation of a variable in the current environment and
can sometimes be convenient.

Wouldn't it be nice to store the figure itself in the repo and
somehow link it to the iris data? This can be done exploiting the
*attach* method, which can store any file in the repo as is,
together with usual annotations (description, tags,
etc.). Internally, *attach* calls *put*, so it accepts most of its
parameters. However, two differences are worth noting:

* **filepath** Instead of an identifier, *attach* takes a file name
(with path). The file name will be also the item identifier.

* **to** This optional parameter tells repo which item the new one
is attached to.


```{r}
fpath <- file.path(repo$root(), "iris2D.pdf")
pdf(fpath)
plot(iris2d, main="2D visualization of the Iris dataset",
     col=repo$get("irisLabels"))
invisible(dev.off())
repo$attach(fpath, "Iris 2D visualization obtained with PCA.",
            c("visualization", "iris", "repodemo"), src, replace=T, to="myiris")
```


*attach* actually creates a copy of the attached file in the repo
root, so the original one can be safely removed:


```{r}
invisible(file.remove(fpath))
```

Wanna see something even nicer? The attached PDF can be opened
using an external PDF viewer directly from repo through the *sys*
method. On a Linux system, this command will run the Evince
document viewer and show *iris2D.pdf*:


```{r, eval=FALSE}
repo$sys("iris2D.pdf", "evince")
```

Back to data analysis, the PCA eigenvalues showed below can give
hints on the reliability of the 2D plot:


```{r}
plot(irispca)
```

 So attaching the plot of the PCA eigenvalues to the plot of the 2D
visualization could be a good idea. This is attaching to an
attachment and is OK.


```{r}
fpath <- file.path(repo$root(), "irispca.pdf")
pdf(fpath)
plot(irispca)
invisible(dev.off())
repo$attach(fpath, "Variance explained by the PCs of the Iris dataset",
            c("visualization", "iris", "repodemo"), src, replace=T, to="iris2D.pdf")
invisible(file.remove(fpath))
```

### Storing results

The following code makes a clustering of the Iris data and stores
it in the repo. There is still one parameter to note here:

* **depends** This tells repo that, in order to compute the *kiris*
variable containing the clustering, data previously stored will be
necessary, namely *myiris*. This dependency annotation is not
mandatory, however it helps to keep things sorted as the repo
grows. Literature on Reproducible Research agrees on the importance
of "provenance traces": setting this value is definitely encouraged
practice.


```{r}
kiris <- kmeans(myiris, 5)$cluster
repo$put(kiris, "iris_5clu", "Kmeans clustering of the Iris data, k=5.",
         c("metadata", "iris", "kmeans", "clustering", "repodemo"), src,
         depends="myiris", T)
```

The following shows what the clustering looks like. The figure
will be attached to the repo as well. Note the dependency.


```{r}
plot(iris2d, main="Iris dataset kmeans clustering", col=kiris)
```

```{r}
fpath <- file.path(repo$root(), "iris2Dkm.pdf")
pdf(fpath)
plot(iris2d, main="Iris dataset kmeans clustering", col=kiris)
invisible(dev.off())
repo$attach(fpath, "Iris K-means clustering.",
            c("visualization", "iris", "clustering", "kmeans", "repodemo"), src,
            replace=T, to="iris_5clu")
invisible(file.remove(fpath))
```


Finally, the last part of the analysis computes a contingency table
of the classes versus clusters. Let's assume that this analysis is
meant to be qualitatively, so the table is not really
important. However it could come handy in the future. This is a
possible use case for the special tag *hide*, which prevents an
item from being shown by default. Attachments are automatically
*hide*-tagged.



```{r}
res <- table(repo$get("irisLabels"), kiris)
repo$put(res, "iris_cluVsSpecies",
         paste("Contingency table of the kmeans clustering versus the",
               "original labels of the Iris dataset."),
         c("result", "iris","validation", "clustering", "repodemo", "hide"),
         src, c("myiris", "irisLabels", "iris_5clu"), T)
```

## Looking at the repository

The *info* method provides details about repo items, but when
called without arguments it shows some self-explaining information
about the repo:



```{r}
repo$info()
```

The repo library supports an S3 *print* method that shows the
contents of the repo. All non-hidden items will be shown, together
with some details, which by defaults are: name, dimensions, size.



```{r}
repo ## by default resolves to print(repo)
```


As expected, hidden items are not shown. The following makes all
the items appear:



```{r}
print(repo, all=T)
```

Items can also be listed selectively. With the following call, only
items tagged with "clustering" will be shown:


```{r}
print(repo, tags="clustering", all=T)
```

All attachments have the *attachment* tag, so they can be
selectively visualized this way:



```{r}
print(repo, tags="attachment", all=T)
```

For the same reasons, one may want to show only hidden items:


```{r}
print(repo, tags="hide", all=T)
```

Finally, also the columns shown can be selected. This command shows
only the tags:


```{r}
repo$print(show="t", all=T)
```

### Showing dependencies

In the previous sections dependencies were set for some items. In
addition, some items were stored as attachment to other
items. Finally, all items were annotated as being produced by
this script. In repo, all these relations are summarized in the
*dependency graph*. The formal representation of the graph is a
matrix, in which the entry (i,j) represent a relation from i to j
of type 1, 2 or 3 (*dependency*, *attachment* or
*generation*). Here's how it looks like:


```{r}
depgraph <- repo$dependencies(plot=F)
rownames(depgraph) <- colnames(depgraph) <- basename(rownames(depgraph))
library(knitr)
kable(depgraph)
```

Omitting the *plot=F* parameter, the *dependencies* method will
plot the dependency graph. This plot requires the *igraph* library.



```{r}
repo$dependencies()
```

This is a small repo and all resources were created by the same
script, so the "generated" edges are not interesting. The three
types of edges can be shown selectively, so here's how the graph
looks like without the "generated" edges:



```{r}
repo$dependencies(generated=F)
```

## Accessing items in the repo

Of course, the only purpose of storing things is to retrieve them
later, and this is why the most used command in repo is *get*. It's
meaning is very straightforward. The following command retrieves
the variable *myiris* from the repo and loads it into the variable
*x* in the current environment.



```{r}
x <- repo$get("myiris")
```

To get additional information about the entry, the *info* command
can be used this way:


```{r}
repo$info("myiris")
```

## Item versions and temporary items

There are actually 4 different ways of adding an object to the
repo:

* Add a new object (*repo$put*)
* Overwrite an existing object (*repo$put(replace=T)*)
* Add a new version of an existing object (*repo$put(addversion = T)*)
* Stash an object as temporary (*repo$stash*)

The first two were covered before, this section will deal with the
other other two.

### Versioning

If you're into Machine Learning, you know that if you run K-means
twice, you're likely going to get two different solutions. Suppose
we want to store such an alternative solution as an additional
version of the *iris_5clu* item. This can be done as follows:


```{r}
kiris2 <- kmeans(myiris, 5)$cluster
repo$put(kiris, "iris_5clu",
         "Kmeans clustering of the Iris data, k=5. Today's version!",
         c("metadata", "iris", "kmeans", "clustering", "repodemo"), src,
         depends="myiris", addversion=T)
```

* **addversion** when TRUE, tells repo to add a new version of an
existing object. The new object will replace the old one and the
old one will be renamed adding the suffix "#N", with N being an
incremental integer.

The new repo looks like the old one:


```{r}
repo
```

Except that *iris_5clu* is actually the one just put (look at the
description):


```{r}
repo$info("iris_5clu")
```

while the old one has been renamed and hidden:


```{r}
repo$print(all=T)
```

However, it can be referred to as any other repo item:


```{r}
repo$info("iris_5clu#1")
```

### Stashing

Repo tries to force the user into building a structured and
annotated repository. However, this implies a small overhead that in
some cases may not be justified. This is when stashing comes
handy.

Consider the case of caching intermediate results. Intermediate
results are not going to be used directly, however they will save
time in case the final results have to be generated again. In such
cases one can just store the intermediate results without
specifying annotations: in repo, this is called *stashing*.

Below is a fake computation that takes 10 secs and that one does
not want to perform during, for example, report generation. One may
set the *dorun* variable to FALSE (BTW that's my personal practice)
so that the script will *get* the precomputed variable from the
repo.

```{r include=FALSE}
dorun <- FALSE
result <- "This took 10 seconds to compute"
repo$stash("result")
```

```{r}
if(dorun) {
    Sys.sleep(10)
    result <- "This took 10 seconds to compute"
    repo$stash("result")
} else result <- repo$get("result")
```

The *stash* function has a rather rough behavior: it will search
the object *name* in the caller environment, create some generic
descriptions and tags, *put* the object into the repo overwriting
stashed items by the same name, and finally hide the newly created
item.


```{r}
repo$info("result")
```

## Handlers

It's a shame that the auto-completion feature by your favorite
editor can not be used on repo item names. Except it can. The
*handlers* method returns a list of functions by the same names of
the items in the repo. Each of these functions can call repo
methods (*get* by default) on the corresponding items. Easier by
examples:


```{r}
h <- repo$handlers()
names(h)
```

Handlers call *get* by default:


```{r}
print(h$iris_cluVsSpecies())
```

The *tag* method (not yet described)  adds a tag to an item:


```{r}
h$iris_cluVsSpecies("tag", "onenewtag")
h$iris_cluVsSpecies("info")
```

One may want to open a repo directly with:


```{r}
h <- repo_open(repo$root())$handlers()
```


In that case, the handler to the repo itself will come handy:


```{r}
h$repo
```

If items are removed or added, handlers may need a refresh:


```{r}
h <- h$repo$handlers()
```

## Clearing the repository

The following R command will completely remove the repo root and all
the objects stored in it (in fact this could be rather dangerous):

```{r}
unlink(repo$root(), recursive=TRUE)
```


## Other features

The repo manual starts at:


```{r, eval=FALSE}
help(repo)
```

All repo methods are also defined as functions in the global
environment. Any call like *repo$func(x)* can be executed as
*repo_func(repo, x)*. In order to get help on the function "func",
try the following:


```{r, eval=FALSE}
help(repo_func)
```

Here's a summary of most repo functions:


|Method     |Description                                                 |
|:----------|:-----------------------------------------------------------|
|rm         |remove an item                                              |
|export     |save an item's object to a regular RDS file                 |
|set        |edit an existing entry                                      |
|tag        |add a tag to an existing entry                              |
|untag      |remove a tag from an existing entry                         |
|sys        |execute a system command on the file referred to by an item |
|tags       |show the complete list of tags in the repo                  |
|check      |check repository integrity                                  |
|stashclear |remove all stashed items                                    |
|root       |get the repo root path                                      |


